---
title: '#USL_P1_477001'
output: html_document
date: "2025-12-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# USL Project 1 
Expected output is to do a choose athe most suitable clustering method, do the cluestering and then forecasting. 

```{r setup, include=FALSE}
# Install missing library on local machine
install.packages("NbClust")
# Load other necessary and already installed libraries
library(tidyverse)
library(cluster)    # For PAM and basic clustering
library(factoextra) # For clustering visualizations
library(NbClust)    # For determining the optimal number of clusters
```

# 1. Data Discrovery 

Before diving into clustering, I need to decide what do I want to do with the data and as well to understand to them. The key idea is to compare historical year with the latest available year (meaning 2024). 

```
## 1.1 Data Inventory: 1990s to 2024

We are checking the availability of data for our key energy variables: Coal, Oil, Gas, Nuclear, and Renewables. We filter for entities with a "Code" to ensure we are looking at individual countries, not aggregated regions like "World" or "Europe".

```{r load_and_inspect}
# Load the raw dataset
df_raw <- read.csv("per-capita-energy-stacked.csv")

# Display the internal structure of the dataframe
# This shows us which variables are numeric and which are characters
str(df_raw)

# Show the first 10 rows for a manual quality check
head(df_raw, 10)
```
